
---
title: "OUwie 2.0"
author: "Jeremy M. Beaulieu and Brian C. O'Meara"
output:
  pdf_document:
    fig_caption: yes
vignette: >
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteIndexEntry{Running hisse}
   \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
require(OUwie)
```

It's been quite a while since we've done any active development of `OUwie`. It was recently brought to our attention that the likelihoods between `OUwie` and `ouch` are not the same when the model assumes different OU means (what we refer to here as "OUM"). We would like to thank and credit Clay Cressler for working through our code to identify the specific causes. In doing so, we have taken this opportunity to release a new version of `OUwie` that provides a cleaner output and provides a number of new capabilities. Some of these new functions were implemented at the request of users, for our own research, and as part of tutorials for various workshops. 

## Bug fixes and deprecated functions

The differences between `OUwie` and `ouch` were traced two issues. There first, being a bug in the most recent version of the weight matrix generation code. For some reason, while looping over the different regimes, the function was not resetting the regime to the regime at the root, and the calculation for the weights for the root regime was missing a $W_{0,i}+e^{(-\alpha_{0,i})}$ term. These are now fixed, and fortunately does not cause any detectable effect on the likelihood, though it does have a slight impact the estimates of the regime optima. 

The second issue is significant and appears to be with the way `OUwie` and `ouch` construct the variance-covariance matrix when assuming stationarity at the root. In Beaulieu et al. (2012), we expanded upon equation A5 in Butler and King (2004), which assumed that the root optima was estimated. We found that the root regime was hard, if not impossible, to estimate with even moderate $\alpha$ parameter (the pull). As a result, by default we drop the root optima and absorb the weight into whatever regime the root was painted. We incorrectly referred to this as "stationarity". As Ho and Ane (2013) proposed that in order to assume true stationarity, the covariance requires an additional variance term to account for the fact that, up until $T=0$, the time at which the clade of interest came into existence, the lineage is assumed to have been evolving in the ancestral regime. For example, consider two tips that diverge at the root. Under the original formulation of Butler and King (2004) and Beaulieu et al. (2012) the covariance between the two tips was zero because $s_{ij}=0$, so $1-e^{-2\alpha s_{ij}}=0$). With the Ho and Ane (2013) method, an additional $V_{ij} = \frac{\sigma^2}{2\alpha}e^{-2\alpha}$ is added to the variance-covariance matrix.

It is worth pointing out, however, that Ho and Ane (2014) cautioned that when there are several regimes the stationary distribution does not have a clear definition. This is because it is a weighted average of the regimes the lineage switched between starting from the origin of life to the most recent common ancestor of the focal clade under study. In other words, it's not as straightforward as integrating over some distribution. For now, `OUwie` assumes a fixed root, where the $\theta_0$ is equal to whatever regime the root is painted (see below on behavior). In order to make OU1 and OUM models consistent with `ouch` we have added the stationarity assumption of Ho and Ane (2014):

```{r, eval=TRUE}
data(tworegime)
pp <- OUwie(tree, trait, model="OUM", root.station=TRUE, scaleHeight=1, shift.point=0)
pp
```

compare this to the new default:

```{r, eval=TRUE}
data(tworegime)
pp <- OUwie(tree, trait, model="OUM", root.station=FALSE, scaleHeight=1, shift.point=0)
pp
```

Note that we have also added the new option `shift.point=` into the function call. This allows users to alter the assumption of where the regime shifts occur. By default `OUwie()` assumes any regime shift occurs halfway down a branch (i.e., `shift.point=0.5`), whereas `ouch` assumes a regime shift occurs at the end of the branch (i.e, `shift.point=0`). Generally speaking, this will have a slight effect on estimates of $\theta_i$ because the position of regime shift point will alter the time spent in each regime.

Finally, we note that we have removed two functions previously available: `OUwie.joint()` and `OUwie.slice()`. The `OUwie.joint()` was developed for a particular study question (see Leslie et al. 2014), otherwise it is not particularly useful. The function `OUwie.slice()` was developed and moderately tested, but it does not seem to work particularly well. Both functions are still available by request, and future work will focus on improving and understanding the behavior of `OUwie.slice()`.


## New Features

## Contour plots

Confidence intervals are usually estimated for individual parameters. However, it is possible for the confidence intervals for pairs of parameters to be much wider than the univariate confidence intervals if the parameters are correlated with one another. For example, for OU models a decrease in σ2 has a similar effect as an increase in α: less variation at the tips. There are differences, so the parameters are identifiable (greater α tends to erase phylogenetic signal, whereas lower σ2 does not). For practical problems, it is possible to have a ridge of nearly equal likelihood where if one changes just one parameter, one moves off the ridge, but where changing both allows one to slide along the ridge. Univariate confidence intervals would not capture this effect. We investigated this by creating contour maps of the likelihood surface for pairs of parameters in our best‐fitting model, letting the other parameters vary to find their maximum likelihood estimates given the values of our focal parameters. Results are shown in Figure 3. For this empirical dataset, parameters appear distinguishable. However, this is an issue that users of this method should consider: even if univariate confidence intervals suggest precise estimates, it is worth considering the effect of varying pairs or potentially correlated parameters.

SHOW THE SIM RESULTS HERE.


### Idenfiability tests$

Ho and Ane (2014) demonstrated that certain regime paintings can produce a ridge in the likelihood surface, which will lead to convergence issues. Specifically, when each regime forms a connected subtree this produces a $m-1$ regime shifts, which is the minimal number . In these situations, the selective optima may not be separately identifiable. In this version of `OUwie` we have implemented a identifiability of the selective regimes check, as both part of the `OUwie()` function and as a standalone function. With `OUwie()` by default `check.identify=TRUE`, and if the check fails, the function will stop. However, this check can be turned off by simply changing `check.identify=FALSE`. The standalone function simply requires the tree with the regimes painted (either as a simmap object or with node labels) and the data set. We will recreate the general scheme shown in Ho and Ane (2014):


SHOW THE SIM SITUATIONS.

```{r, eval=FALSE}
phy <- tree
phy$node.label <- rep(1, length(phy$node.label))
data <- trait
data[,2] <- rep(1, length(data[,1]))

node1 <- getMRCA(phy, c("t1", "t16"))
node2 <- getMRCA(phy, c("t17", "t32"))
node3 <- getMRCA(phy, c("t25", "t32"))


tips.nodes.to.paint1 <- c(node1, getDescendants(tree, node1, curr=NULL))
tips1 <- tips.nodes.to.paint1[tips.nodes.to.paint1<=Ntip(phy)]
nodes1 <- tips.nodes.to.paint1[tips.nodes.to.paint1>Ntip(phy)]

tips.nodes.to.paint2 <- c(node2, getDescendants(tree, node2, curr=NULL))
tips2 <- tips.nodes.to.paint2[tips.nodes.to.paint2<=Ntip(phy)]
nodes2 <- tips.nodes.to.paint2[tips.nodes.to.paint2>Ntip(phy)]

tips.nodes.to.paint3 <- c(node3, getDescendants(tree, node3, curr=NULL))
tips3 <- tips.nodes.to.paint3[tips.nodes.to.paint3<=Ntip(phy)]
nodes3 <- tips.nodes.to.paint3[tips.nodes.to.paint3>Ntip(phy)]

phy$node.label[nodes1-Ntip(phy)] <- 2
phy$node.label[nodes2-Ntip(phy)] <- 3
phy$node.label[nodes3-Ntip(phy)] <- 1

data[tips1,2] <- 2
data[tips2,2] <- 3
data[tips3,2] <- 1


plot(phy)
nodelabels(phy$node.label)
check.identify(phy, data)

```


### OUwie.dredge



### Ancestral trait reconstruction

Since `OUwie` was first released we've received a deluge of requests to allow for ancestral trait reconstruction. One such request made its way onto the public R-SIG-PHYLO discussion forum, which stimulated an important conversation about not whether or not you **could** estimate ancestral states, but, rather, **should** you. The answer is "it's complicated" and we don't recommend it. In our view, the intended use case is just to visualize what the model is saying about evolution to help intuition. For example, is the model something you can believe in? But, if you don't want to listen to us regarding the merits of ancestral trait reconstructions, here is a sample of comments from other experts in the field:

> *So in short, yes, you can do it, with any number of methods. But why? If you can answer your biological question with methods that do not involve estimation of a parameter that is inherently fraught with error, it might be better to go another way. Bottom line - use caution and be thoughtful!* 
-- Marguerite Butler

> *I would add an extra caveat to Marguerite's excellent post: Most researchers work with extant taxa only, ignoring extinction. This causes a massive ascertainment bias, and the character states of the extinct taxa can often be very different to the ancestral state reconstructions, particularly if the evolutionary model is wrong. E.g. there has been an evolutionary trend for example. Ancestral state reconstructions based only on extant taxa should be treated as hypotheses to be tested with fossil data. I wouldn't rely on them for much more.*
-- Simone Blomberg

> *While I am at it, let me echo Simone and Marguerite's warnings. The predicted ancestral states will reflect the process you assumed to predict them. Hence, if you  use them to make inferences about evolution, you will recover your own assumptions. I.e. if you predict from a model with no trend, you will find no trend, etc. Many comparative studies are flawed for this reason.* 
-- Thomas Hansen

> *Let me add more warnings to Marguerite and Thomas's excellent responses. People may be tempted to infer ancestral states and then treat those inferences as data (and also to infer ancestral environments and then treat those inferences as data). In fact, I wonder whether that is not the main use people make of these inferences. But not only are those inferences very noisy, they are correlated with each other. So if you infer the ancestral state for the clade (Old World Monkeys, Apes) and also the ancestral state for the clade (New World Monkeys, (Old World Monkeys, Apes)) the two will typically not only be error-prone, but will also typically be subject to strongly correlated errors.  Using them as data for further inferences is very dubious. It is better to figure out what your hypothesis is and then test it on the data from the tips of the tree, without the intermediate step of taking ancestral state inferences as observations. The popular science press in particular demands a fly-on-the-wall account of what happened in evolution, and giving them the ancestral state inferences as if they were known precisely is a mistake.*
-- Joe Felsenstein

> *The minor twist I would throw in is that it's difficult to make universal generalizations about the quality of ancestral state estimation.  If one is interested in the ancestral state value at node N, it might be reasonably estimated if it is nested high up within the phylogeny, if the rates of change aren't high, etc. And (local) trends etc might well be reliably inferred.  We are pretty confident that the common ancestor of humans and chimps was larger than many deeper primate ancestors, for instance. If N is the root of your available phylogeny, however, you have to be much more cautious.*
-- Nick Matzke

> *I'll also add that I think there's a great deal to be skeptical of ancestral trait reconstruction even when large amounts of fossil data is available. You can try the exercise yourself: simulate pure BM on a non-ultrametric tree with lots of 'extinct' tips, and you'll still find pretty large confidence intervals on the estimates of the trait values. What does it mean to do ancestral trait reconstruction, if our calculations of uncertainty are that broad?*
-- Dave Bapst

These people probably know better than anyone about the power and limitations of the OU model in phylogenetics. So, don't listen to us, listen to them!

Still determined? Ok, fair enough. It is straightforward to run the ancestral trait reconstruction in `OUwie`. All you need is an object of class `OUwie`, which is plugged directly into the new `OUwie.anc()` function:
```{r, eval=FALSE, echo=FALSE}
data(tworegime)
set.seed(42)
ouwiefit <- OUwie(tree, trait, model="OUM", scaleHeight=TRUE, root.station=TRUE, shift.point=0.5, quiet=TRUE)
```

```{r, eval=FALSE, echo=FALSE}
recon <- OUwie.anc(ouwiefit)
```

Ah! If you run the code above a somewhat snarky response is printed to the screen. Since you are here reading this, we think this is sufficient, and you are at least aware that we are not huge fans of this approach. So, there is no need to read the manual (it's basically the same as what you see here), and the proper way to call the `OUwie.anc()` function call is simply add in `knowledge=TRUE`: 

```{r, eval=FALSE, echo=FALSE}
recon <- OUwie.anc(ouwiefit, knowledge=TRUE)
```

From there, you can then easily plot these reconstructions using an internal function that recognizes the `OUwie` and `OUwie.anc` class:

#```{r, fig.align="center", fig.width=6, fig.height=6}
#plot(recon, fsize=0.5)
#```

## Future development

Trendy model.


# References

Beaulieu J.M., Jhwueng D.C., Boettiger C., and O'Meara B.C. 2012. Modeling stabilizing selection: Expanding the Ornstein-Uhlenbeck model of adaptive evolution. Evolution 66:2369-2383.

Butler M.A., King A.A. 2004. Phylogenetic comparative analysis: A modeling approach for adaptive evolution. American Naturalist 164:683-695.

Ho, L.S.T., and C. Ane. 2013. Asymptotic theory with hierarchical autocorrelation: Orstein-Uhlenbeck tree models. The Annals of Statistics, 41: 957-981.

Ho, L.S.T., and C. Ane. 2014. Intrinsic inference difficulties for trait evolution with Ornstein-Uhlenbeck models. Methods in Ecology and Evolution, 5: 1133-1146.

