---
title: "Parameter agreement between hOUwie and fitmultiOU"
subtitle: "The same simulated evolutionary process analyzed two ways"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "90%"
)

report_dir <- normalizePath(getwd(), mustWork = TRUE)
if(!dir.exists(file.path(report_dir, "results"))){
  candidate <- file.path(report_dir, "tools", "benchmarks",
                         "houwie-fitmultiou")
  if(dir.exists(candidate)) report_dir <- normalizePath(candidate)
}
results_root <- file.path(report_dir, "results")

read_profile_results <- function(filename){
  if(!dir.exists(results_root)) return(data.frame())
  profile_dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  paths <- file.path(profile_dirs, filename)
  paths <- paths[file.exists(paths)]
  if(!length(paths)) return(data.frame())
  frames <- lapply(paths, read.csv, stringsAsFactors = FALSE)
  columns <- unique(unlist(lapply(frames, names)))
  frames <- lapply(frames, function(x){
    missing <- setdiff(columns, names(x))
    for(column in missing) x[[column]] <- NA
    x[columns]
  })
  do.call(rbind, frames)
}

successful <- function(x){
  if(!nrow(x) || !"error" %in% names(x)) return(x)
  x[is.na(x$error) | x$error == "", , drop = FALSE]
}

fits <- successful(read_profile_results("fits.csv"))
evaluations <- successful(read_profile_results("evaluations.csv"))
pilot_fits <- if(nrow(fits)) fits[fits$profile == "pilot", , drop = FALSE] else
  data.frame()
article_fits <- if(nrow(fits)) fits[fits$profile == "article", , drop = FALSE] else
  data.frame()

candidate_results_path <- file.path(
  results_root, "parameter-comparison", "candidate-likelihoods.csv"
)
candidate_results <- if(file.exists(candidate_results_path)) {
  successful(read.csv(candidate_results_path, stringsAsFactors = FALSE))
} else data.frame()

fixed_history_results_path <- file.path(
  results_root, "parameter-comparison",
  "fixed-history-candidate-likelihoods.csv"
)
fixed_history_results <- if(file.exists(fixed_history_results_path)) {
  successful(read.csv(fixed_history_results_path, stringsAsFactors = FALSE))
} else data.frame()

known_history_results_path <- file.path(
  results_root, "parameter-comparison",
  "known-history-continuous-likelihoods.csv"
)
known_history_results <- if(file.exists(known_history_results_path)) {
  read.csv(known_history_results_path, stringsAsFactors = FALSE)
} else data.frame()

diagnosis_root <- file.path(results_root, "history-integration-diagnosis")
read_diagnosis <- function(filename){
  path <- file.path(diagnosis_root, filename)
  if(file.exists(path)) successful(read.csv(path, stringsAsFactors = FALSE)) else
    data.frame()
}
sampling_modes <- read_diagnosis("sampling-modes.csv")
history_counts <- read_diagnosis("history-counts.csv")
within_edge <- read_diagnosis("within-edge-resolution.csv")
small_exact <- read_diagnosis("small-tree-exact-enumeration.csv")
small_samples <- read_diagnosis("small-tree-sampled-subsets.csv")
importance_results <- read_diagnosis("importance-sampling.csv")
root_treatments <- read_diagnosis("fitmultiou-root-treatments.csv")

recovery_path <- file.path(
  results_root, "parameter-recovery-suite", "fits.csv"
)
recovery_suite <- if(file.exists(recovery_path)) {
  successful(read.csv(recovery_path, stringsAsFactors = FALSE))
} else data.frame()

truth <- c(theta_a = 5, theta_b = 10, alpha = 0.5,
           sigma_sq = 0.2, q = 0.3)
parameter_columns <- names(truth)
parameter_labels <- c(
  theta_a = "Theta a",
  theta_b = "Theta b",
  alpha = "Alpha",
  sigma_sq = "Sigma-squared",
  q = "q"
)

post_parameters <- data.frame(
  Source = c(
    "Generating values",
    "fitmultiOU estimate in post",
    "hOUwie estimate in post"
  ),
  theta_a = c(5, 5.0209, 4.4566072),
  theta_b = c(10, 10.1985, 88.5531307),
  alpha = c(0.5, 0.4919, 0.0123027),
  sigma_sq = c(0.2, 0.1839, 0.7764039),
  q = c(0.3, 0.347199, 0.1187636),
  stringsAsFactors = FALSE
)

format_range <- function(x, digits = 3){
  sprintf(
    paste0("%.", digits, "f [%.", digits, "f, %.", digits, "f]"),
    median(x), min(x), max(x)
  )
}

pilot_parameter_table <- function(x){
  result <- data.frame(
    Source = "Generating values",
    `Theta a` = sprintf("%.3f", truth["theta_a"]),
    `Theta b` = sprintf("%.3f", truth["theta_b"]),
    Alpha = sprintf("%.3f", truth["alpha"]),
    `Sigma-squared` = sprintf("%.3f", truth["sigma_sq"]),
    q = sprintf("%.3f", truth["q"]),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  for(method in c("fitmultiOU", "hOUwie")){
    z <- x[x$method == method, , drop = FALSE]
    if(!nrow(z)) next
    result <- rbind(result, data.frame(
      Source = paste0(method, " median [range]"),
      `Theta a` = format_range(z$theta_a),
      `Theta b` = format_range(z$theta_b),
      Alpha = format_range(z$alpha),
      `Sigma-squared` = format_range(z$sigma_sq),
      q = format_range(z$q),
      check.names = FALSE,
      stringsAsFactors = FALSE
    ))
  }
  result
}

parameter_stability_table <- function(x){
  result <- do.call(rbind, lapply(split(x, x$method), function(z){
    data.frame(
      Method = z$method[1],
      `Theta a SD` = sd(z$theta_a),
      `Theta b SD` = sd(z$theta_b),
      `Alpha SD` = sd(z$alpha),
      `Sigma-squared SD` = sd(z$sigma_sq),
      `q SD` = sd(z$q),
      check.names = FALSE,
      row.names = NULL
    )
  }))
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 4)
  result
}

candidate_summary <- function(x){
  if(!nrow(x)) return(data.frame())
  x$delta_from_generating <- NA_real_
  for(method in unique(x$method)){
    method_rows <- which(x$method == method)
    method_data <- x[method_rows, , drop = FALSE]
    reference <- method_data[
      method_data$candidate == "Generating parameters",
      c("repetition", "loglik")
    ]
    names(reference)[2] <- "reference_loglik"
    matched <- merge(
      method_data[, c("repetition", "candidate")],
      reference,
      by = "repetition",
      all.x = TRUE,
      sort = FALSE
    )
    key <- paste(method_data$repetition, method_data$candidate)
    matched_key <- paste(matched$repetition, matched$candidate)
    x$delta_from_generating[method_rows] <-
      method_data$loglik - matched$reference_loglik[match(key, matched_key)]
  }
  groups <- split(x, interaction(x$method, x$candidate, drop = TRUE))
  result <- do.call(rbind, lapply(groups, function(z){
    data.frame(
      Method = z$method[1],
      Candidate = z$candidate[1],
      Repetitions = nrow(z),
      `Mean log likelihood` = mean(z$loglik),
      `Log-likelihood SD` = if(nrow(z) > 1) sd(z$loglik) else 0,
      `Median difference from generating values` =
        median(z$delta_from_generating),
      check.names = FALSE,
      row.names = NULL
    )
  }))
  candidate_order <- match(result$Candidate, c(
    "Generating parameters",
    "fitmultiOU estimate from post",
    "hOUwie estimate from post"
  ))
  result <- result[order(result$Method, candidate_order), ]
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 3)
  result
}

fixed_history_table <- function(x){
  if(!nrow(x)) return(data.frame())
  x$difference_from_generating <- NA_real_
  for(bank in unique(x$bank_source)){
    rows <- which(x$bank_source == bank)
    reference <- x$loglik[
      rows[x$candidate[rows] == "Generating parameters"]
    ]
    x$difference_from_generating[rows] <- x$loglik[rows] - reference
  }
  result <- x[, c(
    "bank_source", "candidate", "loglik", "difference_from_generating"
  )]
  names(result) <- c(
    "Histories sampled under", "Candidate", "Log likelihood",
    "Difference from generating values"
  )
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 3)
  result
}

known_history_table <- function(x){
  if(!nrow(x)) return(data.frame())
  reference <- x$continuous_loglik[
    x$candidate == "Generating parameters"
  ]
  result <- data.frame(
    Candidate = x$candidate,
    `Continuous log likelihood` = round(x$continuous_loglik, 3),
    `Difference from generating values` = round(
      x$continuous_loglik - reference, 3
    ),
    check.names = FALSE
  )
  result
}

candidate_contrasts <- function(x, value_column, setting_column){
  if(!nrow(x)) return(data.frame())
  settings <- unique(x[[setting_column]])
  result <- do.call(rbind, lapply(settings, function(setting){
    z <- x[x[[setting_column]] == setting, , drop = FALSE]
    generating <- z[z$candidate == "Generating parameters",
                    c("repetition", value_column), drop = FALSE]
    names(generating)[2] <- "generating"
    merged <- merge(z, generating, by = "repetition", all.x = TRUE)
    merged$difference <- merged[[value_column]] - merged$generating
    groups <- split(merged, merged$candidate)
    do.call(rbind, lapply(groups, function(group){
      data.frame(
        Setting = setting,
        Candidate = group$candidate[1],
        `Mean log likelihood` = mean(group[[value_column]]),
        `Mean difference from generating values` = mean(group$difference),
        `Difference SD` = sd(group$difference),
        check.names = FALSE,
        row.names = NULL
      )
    }))
  }))
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 3)
  result
}

small_error_table <- function(x){
  if(!nrow(x)) return(data.frame())
  groups <- split(x, interaction(x$n_sim, x$candidate, drop = TRUE))
  result <- do.call(rbind, lapply(groups, function(z){
    data.frame(
      Histories.requested = z$n_sim[1],
      Candidate = z$candidate[1],
      Histories.retained = round(mean(z$retained_histories), 1),
      Mean.error = mean(z$error_from_exact),
      Worst.error = min(z$error_from_exact),
      row.names = NULL
    )
  }))
  names(result) <- c(
    "Histories requested", "Candidate", "Histories retained",
    "Mean error from exact likelihood", "Worst error"
  )
  result <- result[order(result$`Histories requested`, result$Candidate), ]
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 3)
  result
}

recovery_summary_table <- function(x){
  if(!nrow(x)) return(data.frame())
  x <- x[x$method == "hOUwie", , drop = FALSE]
  x$theta_rmse <- sqrt(
    ((x$theta_a - x$true_theta_a)^2 +
       (x$theta_b - x$true_theta_b)^2) / 2
  )
  x$separation_ratio <-
    (x$theta_b - x$theta_a) / (x$true_theta_b - x$true_theta_a)
  x$alpha_ratio <- x$alpha / x$true_alpha
  x$sigma_ratio <- x$sigma_sq / x$true_sigma_sq
  x$q_ratio <- x$q / x$true_q
  groups <- split(x, x$scenario)
  result <- do.call(rbind, lapply(groups, function(z){
    data.frame(
      Scenario = z$scenario[1],
      `Theta RMSE` = median(z$theta_rmse),
      `Optimum separation ratio` = median(z$separation_ratio),
      `Alpha ratio` = median(z$alpha_ratio),
      `Sigma-squared ratio` = median(z$sigma_ratio),
      `q ratio` = median(z$q_ratio),
      check.names = FALSE,
      row.names = NULL
    )
  }))
  scenario_order <- c(
    "baseline", "weak-alpha", "strong-alpha", "low-sigma-squared",
    "high-sigma-squared", "small-optimum-separation",
    "large-optimum-separation", "low-q", "high-q"
  )
  result <- result[match(scenario_order, result$Scenario), ]
  numeric <- vapply(result, is.numeric, logical(1))
  result[numeric] <- lapply(result[numeric], round, 3)
  result
}
```

The parameter estimates are not similar on Liam's 300-tip simulation.
`fitmultiOU()` stays close to the values that generated the data. hOUwie favors
a region with much weaker attraction, a larger second optimum, more diffusion,
and a lower transition rate. That difference remains after the recent hOUwie
performance and stability changes.

Both programs fit a discrete-character-dependent OUM process. hOUwie averages
over sampled character histories, while `fitmultiOU()` represents the joint
process on a finite grid. The comparisons below use the exact tree and traits
reconstructed from Liam's seed.

## The parameter estimates

The simulation used $\theta_a=5$, $\theta_b=10$, $\alpha=0.5$,
$\sigma^2=0.2$, and $q=0.3$.

```{r post-parameters}
display_post <- post_parameters
display_post[-1] <- lapply(display_post[-1], function(x) round(x, 4))
names(display_post) <- c(
  "Source", "Theta a", "Theta b", "Alpha", "Sigma-squared", "q"
)
knitr::kable(display_post, row.names = FALSE)
```

The `fitmultiOU()` estimate differs from the generating values by 0.02 for
$\theta_a$, 0.20 for $\theta_b$, 0.008 for $\alpha$, 0.016 for $\sigma^2$, and
0.047 for $q$. The hOUwie estimate has $\theta_b=88.55$, $\alpha=0.012$,
$\sigma^2=0.776$, and $q=0.119$. These are different parameter regions, not
minor numerical variations around the same solution.

## Evaluating the same parameter values again

The current hOUwie likelihood was evaluated at three parameter vectors:

1. the generating values;
2. the `fitmultiOU()` estimates printed in the post; and
3. the hOUwie estimates printed in the post.

Each hOUwie value uses 100 histories. Ten paired random streams repeat the
comparison, with the same stream used for all three candidates in each repeat.
The `fitmultiOU()` calculation uses a deterministic 100-level grid.

```{r candidate-results, results='asis'}
if(nrow(candidate_results)){
  print(knitr::kable(candidate_summary(candidate_results), row.names = FALSE))
}else{
  cat("Run `compare-parameters.R` to populate this section.\n")
}
```

```{r candidate-interpretation, results='asis'}
if(nrow(candidate_results)){
  summary <- candidate_summary(candidate_results)
  houwie <- summary[summary$Method == "hOUwie", ]
  fitmulti <- summary[summary$Method == "fitmultiOU", ]
  best_houwie <- houwie$Candidate[which.max(houwie$`Mean log likelihood`)]
  best_fitmulti <- fitmulti$Candidate[which.max(fitmulti$`Mean log likelihood`)]
  cat("hOUwie gives its highest mean value to **", best_houwie,
      "**. `fitmultiOU()` gives its highest value to **", best_fitmulti,
      "**. hOUwie places the low-alpha/high-theta vector about 406 ",
      "log-likelihood units above the generating values on the median paired ",
      "repeat. `fitmultiOU()` places that vector 113 log-likelihood units ",
      "below the generating values.\n", sep = "")
}
```

The parameter translation was checked directly. hOUwie receives the values as
$q$, $\alpha$, $\sigma^2$, $\theta_a$, and $\theta_b$; `fitmultiOU()` receives
the same five values in its own documented order and scale.

## Holding the histories constant

hOUwie usually samples a new collection of histories for each parameter vector.
The next comparison removes that moving part. One collection of 100 histories
was sampled under each candidate, then all three parameter vectors were
evaluated on the same collection.

```{r fixed-history-results, results='asis'}
if(nrow(fixed_history_results)){
  print(knitr::kable(
    fixed_history_table(fixed_history_results), row.names = FALSE
  ))
}else{
  cat("Run `compare-parameters.R` to populate this section.\n")
}
```

The ranking is unchanged in all three collections. The hOUwie parameter vector
is 386 to 472 log-likelihood units above the generating values. The
`fitmultiOU()` vector is 65 to 76 units below them. Randomly drawing a different
set of histories is therefore not enough to explain the disagreement.

## The continuous calculation on the known history

Because these data were simulated, the exact character history that generated
the continuous traits is known. Evaluating only the continuous OU likelihood on
that history separates the OU calculation from uncertainty about ancestral
states and transitions.

```{r known-history-results, results='asis'}
if(nrow(known_history_results)){
  print(knitr::kable(
    known_history_table(known_history_results), row.names = FALSE
  ))
}else{
  cat("Run `compare-parameters.R` to populate this section.\n")
}
```

On the known history, hOUwie's continuous calculation ranks the generating
values first. The `fitmultiOU()` estimate is only 2.5 log-likelihood units lower,
whereas the hOUwie estimate from the post is 193.7 units lower. The continuous
OU calculation and parameter order are behaving as expected. The divergence
appears when the unknown discrete histories are combined with the continuous
trait likelihood.

## Sampling histories using the continuous trait

The standard hOUwie call samples node histories using the discrete tip states.
The package can instead use the continuous trait when constructing the node
proposal. The four available combinations were evaluated three times with 100
histories per candidate.

```{r sampling-mode-results, results='asis'}
if(nrow(sampling_modes)){
  mode_table <- candidate_contrasts(
    sampling_modes, "current_loglik", "setting"
  )
  mode_names <- c(
    `discrete-only` = "Discrete only",
    `continuous-informed-nodes` = "Continuous-informed nodes",
    adaptive = "Adaptive search",
    `continuous-informed-and-adaptive` =
      "Continuous-informed nodes plus adaptive search"
  )
  mode_table$Setting <- unname(mode_names[mode_table$Setting])
  print(knitr::kable(mode_table, row.names = FALSE))
}
```

The discrete-only calculation favors the low-alpha hOUwie vector by 432
log-likelihood units on average. Continuous-informed node sampling reverses the
ranking in every repeat. The generating vector is then 204 units above the old
hOUwie vector, and the `fitmultiOU()` vector is only 4.6 units below the
generating values.

Adaptive search by itself is much less stable. Combining it with
continuous-informed nodes again favors the old vector. The likelihood is
therefore responding to the method used to find histories, even though the
underlying target likelihood has not changed.

## More histories and finer histories

```{r history-count-results, results='asis'}
if(nrow(history_counts)){
  print(knitr::kable(
    candidate_contrasts(history_counts, "current_loglik", "setting"),
    row.names = FALSE
  ))
}
```

Increasing the discrete-only sample from 25 to 500 histories narrows the gap,
but the low-alpha vector remains 359 log-likelihood units above the generating
values. Dividing long branches into two-, one-, and half-time-unit intervals
also leaves the low-alpha vector hundreds of units ahead. The problem is not a
small history count alone, nor is it caused by placing a transition halfway
along each branch.

Changing the `fitmultiOU()` root treatment among maximum likelihood, flat, and
nuisance roots also leaves its parameter ranking unchanged. Under all three,
the `fitmultiOU()` estimate is slightly above the generating values and the old
hOUwie estimate is 113 to 115 units below them.

## Exact enumeration on a small tree

A six-tip binary tree has 32 possible assignments of two states to its five
internal nodes. This is small enough to sum every history and calculate the
exact likelihood targeted by hOUwie's midpoint-history model.

```{r exact-small-tree-results, results='asis'}
if(nrow(small_exact)){
  exact_display <- small_exact
  names(exact_display) <- c(
    "Candidate", "Possible histories", "Exact midpoint log likelihood"
  )
  exact_display$`Exact midpoint log likelihood` <- round(
    exact_display$`Exact midpoint log likelihood`, 3
  )
  print(knitr::kable(exact_display, row.names = FALSE))
}
```

The exact calculation ranks the generating parameters first, with the
`fitmultiOU()` vector only 0.03 units lower and the old hOUwie vector 5.95 units
lower.

```{r sampled-small-tree-results, results='asis'}
if(nrow(small_samples)){
  print(knitr::kable(small_error_table(small_samples), row.names = FALSE))
}
```

Requesting 100 histories recovers all 32 unique possibilities and matches the
exact likelihood to numerical precision. With five histories, however, the
generating likelihood is underestimated by 5.89 units on average and by as much
as 28.60 units. The old hOUwie vector is underestimated by only 0.69 units on
average. The sampled calculation therefore penalizes different parameter
vectors by very different amounts.

## Cause of the discrepancy

For a sampled set of histories $H_j$, hOUwie currently calculates

$$
\log\sum_j P(H_j,Y\mid q)P(X\mid H_j,\eta),
$$

where $Y$ is the observed discrete character, $X$ is the continuous trait, and
$\eta$ contains the OU parameters. This is exact when the sum contains every
possible history. With a subset, it is a lower bound whose missing mass depends
on the parameter vector and on the proposal used to find histories.

A 300-tip binary tree has $2^{299}$ possible assignments of two states to its
internal nodes before adding any within-branch states. One hundred or even 500
histories cannot approximate that sum by coverage. The discrete-only proposal
rarely finds histories compatible with strong attraction toward the two true
optima. It more readily finds histories that work with weak attraction and high
diffusion, so the low-alpha region loses much less likelihood mass when the sum
is truncated.

A Monte Carlo likelihood instead requires the proposal probability $r(H_j)$:

$$
\log\left[\frac{1}{n}\sum_j
\frac{P(H_j,Y\mid q)P(X\mid H_j,\eta)}{r(H_j)}\right].
$$

The prototype computes the exact sequential proposal probability and applies
this correction. It confirms the algebra: under the discrete-only proposal,
$\log P(H,Y)-\log r(H)$ is constant to about $10^{-13}$. It also shows why a
simple correction is not yet a practical replacement. At both 100 and 500
histories, the effective sample size for the generating and `fitmultiOU()`
vectors is usually close to one; a single history carries more than 90% of the
weight. Continuous-informed sampling restores the expected ordering, but its
weights are still too concentrated for stable optimization.

The immediate numerical discrepancy is caused by candidate-dependent coverage
of history space. The continuous OU likelihood, parameter translation,
within-branch resolution, and root treatment have each been ruled out. A robust
repair needs a better trait-informed proposal together with explicit importance
weights, or a different integration method such as the joint-state pruning used
by `fitmultiOU()`.

## Parameter recovery beyond Liam's settings

A compact recovery sweep varied attraction strength, diffusion, optimum
separation, and transition rate around Liam's baseline. Each scenario used
three independently simulated 50-tip datasets. To make the test favorable to
hOUwie, every optimization began at the generating parameter vector, used
common random numbers, and retained 50 histories. These were deliberately short
150-evaluation fits: they test whether hOUwie's objective pulls the optimizer
away from truth, rather than claiming to locate a final maximum.

The table reports medians across the three datasets. A ratio of one is the
generating value; a negative optimum-separation ratio means that the two fitted
optima reversed order.

```{r recovery-suite-results, results='asis'}
if(nrow(recovery_suite)){
  print(knitr::kable(
    recovery_summary_table(recovery_suite), row.names = FALSE
  ))
}else{
  cat("Run `parameter-recovery-suite.R` to populate this section.\n")
}
```

Across all 27 hOUwie fits, alpha was below its generating value in 23, q was
below its generating value in 22, sigma-squared was above its generating value
in 19, and the fitted optima reversed order in 10. The baseline median alpha was
28% of truth, median sigma-squared was 2.88 times truth, and median q was 48% of
truth. Strong attraction, low diffusion, large optimum separation, and frequent
state transitions did not rescue recovery. These are precisely the settings in
which a history sampled without enough information from the continuous trait is
most costly, and hOUwie responded by weakening attraction or inflating
diffusion so that more sampled histories could accommodate the observations.

One matched `fitmultiOU()` fit per scenario provides a numerical reference. It
kept the optima near their generating values in all nine scenarios and was close
for most other parameters. As expected from only 50 tips, alpha was weakly
identified when the generating alpha was 0.05, and q was underestimated when
the generating q was 1. Those ordinary finite-sample errors are much smaller and
less systematically directed than the hOUwie deviations.

## Estimates from the smaller matched run

The 75-tip pilot runs both optimizers three times from the same starting values.
It uses 25 histories for hOUwie, 40 grid levels for `fitmultiOU()`, one optimizer
start, and a budget of 200 evaluations or iterations.

```{r pilot-parameters, results='asis'}
if(nrow(pilot_fits)){
  print(knitr::kable(pilot_parameter_table(pilot_fits), row.names = FALSE))
}else{
  cat("Pilot results have not been generated yet.\n")
}
```

The pilot estimates are also not similar. `fitmultiOU()` remains close to the
generating values for theta, alpha, and q, with a lower estimate of
sigma-squared. hOUwie again moves toward a low-alpha region.

The pilot was deliberately stopped early. `fitmultiOU()` reached its iteration
limit, and hOUwie used between 186 and 199 of its 200 allowed evaluations. The
result shows the direction taken under a small numerical budget rather than a
final maximum.

## Stability of the parameter estimates

```{r parameter-stability, results='asis'}
if(nrow(pilot_fits)){
  print(knitr::kable(parameter_stability_table(pilot_fits), row.names = FALSE))
}
```

`fitmultiOU()` returns the same pilot estimates on all three calls because the
data, grid, and starting values are fixed. hOUwie changes across random streams,
with most of the movement in this pilot appearing in $\theta_a$ and $\alpha$.

Increasing the number of sampled histories reduces noise in the hOUwie
likelihood. Across five streams, its standard deviation falls from 31.0 units
with 10 histories to 16.8 with 25 histories and 11.9 with 50 histories.

## Full optimization

The `article` profile preserves Liam's expensive settings: 100 grid levels and
2,000 iterations for `fitmultiOU()`, plus ten hOUwie starts with 100 histories
and up to 1,000 evaluations per start. This is the comparison that can determine
whether the fully optimized upgraded hOUwie estimates match `fitmultiOU()` on
Liam's exact dataset.

```{r article-results, results='asis'}
if(nrow(article_fits)){
  article_table <- article_fits[, c(
    "method", "starts", "cores", "theta_a", "theta_b", "alpha",
    "sigma_sq", "q", "evaluations", "convergence"
  )]
  names(article_table) <- c(
    "Method", "Starts", "Cores", "Theta a", "Theta b", "Alpha",
    "Sigma-squared", "q", "Evaluations", "Convergence"
  )
  numeric <- vapply(article_table, is.numeric, logical(1))
  article_table[numeric] <- lapply(article_table[numeric], round, 4)
  print(knitr::kable(article_table, row.names = FALSE))
}else{
  cat("The full optimization has not been run in this one-core environment. ",
      "Start or resume it with:\n\n")
  cat("```sh\n")
  cat("Rscript tools/benchmarks/houwie-fitmultiou/run-comparison.R article\n")
  cat("```\n")
}
```

The broader `study` profile repeats the fit across three tree sizes and three
independently simulated datasets.

## Timing appendix

On the single-core 300-tip check, evaluating 100 hOUwie histories took a median
1.715 seconds. Evaluating the 100-level `fitmultiOU()` grid took a median 6.055
seconds. These settings do not imply equal numerical accuracy.

The full timing records remain in `results/pilot/` and `results/scaling/`.

## Files

- `compare-parameters.R` evaluates the candidate parameter vectors with new
  histories, shared histories, and the known generating history.

- `diagnose-history-integration.R` runs the history-count, proposal,
  within-edge, root-treatment, and exact-enumeration checks.

- `importance-sampling-prototype.R` calculates explicit proposal probabilities,
  importance weights, and effective sample sizes.

- `parameter-recovery-suite.R` runs the nine-scenario recovery sweep from the
  generating parameters.

- `run-comparison.R` contains the simulation, optimization profiles, result
  extraction, and resume logic.

- `comparison.Rmd` is this editable report.

- `results/` contains the CSV results and exact simulated inputs.

The benchmark requires the development version of *phytools* containing
`fitmultiOU()`. A separate library can be selected without replacing an existing
installation:

```sh
PHYTOOLS_LIB=/path/to/development-library \
  Rscript tools/benchmarks/houwie-fitmultiou/compare-parameters.R
```

Render the report after any run:

```r
rmarkdown::render(
  "tools/benchmarks/houwie-fitmultiou/comparison.Rmd"
)
```
